---
title: "Explore PSA Assay"
author: "Lauren Howard"
date: "`r Sys.Date()`"
output: html_document
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_file = paste0("../Results/IRONMAN PSA exploration.html")) })
---

<!-- # Set defaults -->
```{r, echo =FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning=FALSE, 
                      message=FALSE, 
                      tidy=TRUE,
                      fig.align='center',
                      fig.show='hold',
                      size='footnotesize',
                      fig.width=10,
                      fig.height=6)
```


<!-- # Load Packages -->

```{r}
# load packages
library(readxl)
library(tidyverse)
library(lubridate)
library(here)
library(dplyr)
library(labelled)
library(writexl)
library(stringr)
library(janitor)
# load ironman package
# Repo: https://github.com/pcctc/ironer
# Website: https://pcctc.github.io/ironer/
#devtools::install_github("pcctc/ironer")
library(ironer)
library(DT)

```


<!-- # Read data -->

```{r}
load("V:/IRONMAN/2024-11-04/2024-11-04_ironman.Rdata")

```

<!-- # Site -->

```{r}
subject_data <- medidata$subject %>%
  select(subject, 
         site, 
         site_name_short,
         site_country)
```


<!-- # PSA data -->

```{r}
## PSA
```{r}
# # pull data from PSA and HXPSA
all_psa <- medidata$psa %>%
  select(subject, psa) %>%
  bind_rows(medidata$hxpsa %>%
              select(subject, psa=hxpsaval)) %>%
  # replace <# to 0 and ># to #, replace comma to period
  mutate(# replace <# to #
         psa = gsub("<", "", psa),
         # replace ># to #
         psa = str_remove(psa, ">"),
         # replace comma to period
         psa = gsub(",", "\\.", psa),
         # replace double period to period
         psa = gsub("\\.\\.", "\\.", psa),
         # replace slash to to period
         psa = gsub("/", "\\.", psa),
         # replace .0. to 0.
         psa = gsub("\\.0\\.", "0\\.", psa),
         # remove leading character
         psa = gsub("z", "", psa),
         # strip spaces
         psa = trimws(psa),
         # convert to numeric
         psa = as.numeric(psa)) %>%
  filter(!is.na(psa)) %>% 
  left_join(subject_data, by="subject")

less_psa <- medidata$psa %>%
  select(subject, psa) %>%
  bind_rows(medidata$hxpsa %>%
              select(subject, psa=hxpsaval)) %>%
  filter(grepl("<", psa)) %>% 
    mutate(# remove space after <
        psa = gsub("< ", "<", psa),
      # replace comma to period
         psa = gsub(",", "\\.", psa),
         # replace double period to period
         psa = gsub("\\.\\.", "\\.", psa),
         # replace slash to to period
         psa = gsub("/", "\\.", psa),
         # replace .0. to 0.
         psa = gsub("\\.0\\.", "0\\.", psa),
         # remove leading character
         psa = gsub("z", "", psa),
         # strip spaces
         psa = trimws(psa),
         # remove trailing zeros
         psa = sub("0+$", "", psa)) %>% 
  left_join(subject_data, by="subject") %>% 
  group_by(psa, site) %>% 
  summarize(n = n()) %>% 
  arrange(site)

```

# Count of PSAs with "<" by site
```{r}

datatable(less_psa, 
          editable = TRUE,
          rownames = FALSE,
          extensions = 'Buttons',
          options = list(
            dom = 'Blfrtip')
          )
```


# Lowest PSA by site
```{r}
min_psa_site <- all_psa %>% 
  group_by(site, site_country) %>% 
  summarize(n = n(), min_psa = min(psa)) %>% 
  arrange(-min_psa)

datatable(min_psa_site, 
          editable = TRUE,
          rownames = FALSE,
          extensions = 'Buttons',
          options = list(
            dom = 'Blfrtip')
          )

```

# Distribution of PSAs <=0.2
```{r}
all_psa %>% 
  filter(psa<=0.2) %>% 
  ggplot(aes(x=psa)) +
  geom_histogram(binwidth = 0.01, fill = "blue", color = "black") +
  labs(title = "Histogram of PSAs <=0.02", x = "PSA", y = "Frequency")

```

