---
title: "IRONMAN_Prelim_PSA 6 months 10192024"
author: "Roy Soumyajit"
date: "2024-10-21"
output: html_document
---

```{r setup, include=F, warning=F}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=999)
library(pacman)
p_load(tidyverse, dynpred, ggplot2, survival, survminer, KMunicate, ggsurvfit, tidycmprsk, lme4, readr, readxl)
library(adjustedCurves)
library(WeightIt)
library(SuperLearner)
library(pbkrtest)
library(mice)
library(haven)
library(table1)
library(naniar)
library(finalfit)
library(tableone)
library(lubridate)
library(ironer)
```


## Methods
*We included patients from PCCTC IRONMAN cohort with available baseline data on treatment regimen and PSA values at baseline, and 1 PSA value after baseline and by month 9.*

*We stratified patients into 3 groups depending on the lowest PSA they had at or before 6 months. These groups included PSA nadir <0.02 ng/mL, 0.02 to 0.2 ng/mL, and >0.2 ng/mL, respectively.*

*A landmark population was constructed with patients who were alive and under follow-up for more than 12 months.*

*We determined 3-year OS rates for the three PSA nadir groups. Subsequently we applied Cox proportional hazard regression model to determine the adjusted association of PSA nadir with OS. We used two models - one with time to PSA nadir and one without time to PSA nadir*


*We applied competing risk methods described by Fine and Gray to determine the cumulative incidence of prostate cancer specific mortality (PCSM) in these three PSA nadir groups. Deaths from causes other than prostate cancer were considered as competing events. Again we applied multivariable CRR with and without time to PSA nadir.*


# Data analysis and Cleaning {.tabset}

## Import data and merge

```{r data}

# load("~/Documents/IRONMAN/2024-09-23_ironman.RData")
load("V:/IRONMAN/2024-11-04/2024-11-04_ironman.Rdata")

###########################################################
cohort <- medidata$cohort%>%
  filter(cohort=="mHSPC")%>%
  select(subject, cohort, denovomet_disease)%>%
  distinct(subject, .keep_all = T)

```


## Assigning first line treatment
```{r data treatment, echo=T}

### Treatment preparation
treatment_prep <- 
  medidata$ca_cm_derived %>% 
  dplyr::select(
    subject, treatment, exstdat_int, exendat_int, 
    treatment_recode = trt_treatment_recode, treatment_category = trt_treatment_category)

treatment_prep <- treatment_prep%>%
  filter(subject %in% cohort$subject)

treatment_prep <- treatment_prep%>%
  filter(!is.na(exstdat_int)) %>% 
  # remove treatments not counted as a line of therapy
  filter(!treatment_category %in% c("Steroid", "Bone-Modifying Agent", "Bisphosphonates", "Antidiabetic"))%>%
  # keep relevant fields
  select(subject, exstdat_int, exendat_int, treatment_category, treatment_recode)


##### Collecting consent date and matching with the mHSPC cohort
consent <- medidata$ic%>%select(subject, cnstdate_int)
consent <- consent%>%
  filter(subject %in% cohort$subject)

#### To calculate gap from consent to start of treatment

treatment_prep <- left_join(treatment_prep, consent, by = "subject")

treatment_prep <- treatment_prep%>%
  mutate(gap = as.duration(cnstdate_int%--%exstdat_int)/ddays(1))%>%
  filter(gap>=(-90))

#### Next step
treatment_lot <- treatment_prep %>%
  # sort by subject and treatment start date
  arrange(subject, exstdat_int) %>%
  # within each subject, assign line of therapy
  group_by(subject) %>%
  group_modify(~ ironer::assign_lot(
    .data = .x, 
    # note, you could use treatment_category here as well to get a high level
    # view of line of therapy by category
    treatment = treatment_category,
    dt_trt_observed = exstdat_int, 
    dt_treatment_end = exendat_int
  ))%>%
  ungroup()  

```

## Selecting Patients who Received ARPI
```{r}

############################### First lot ARPI
first_lot <- treatment_lot%>%
  filter(lot==1)
  

first_arpi <- first_lot%>%
  filter(str_detect(treatment_category,"AR Signaling Inhibitor"))%>%
  distinct(subject, .keep_all = T)


first_arpi <- first_arpi%>%
  select(subject,regimen,treatment_category,treatment_recode)

################################ Second lot ARPI

second_lot <- treatment_lot%>%
  filter(lot==2)

### First select who received ADT in first lot
first_adt <- first_lot%>%
  filter(treatment_category=="ADT" & treatment_recode != "Bicalutamide")%>%
  distinct(subject, .keep_all = T)%>%
  rename(first_treatment_category = treatment_category, 
         first_dt_lot_start = dt_lot_start,
         first_dt_lot_last_obs = dt_lot_last_obs)%>%
  select(subject,first_treatment_category,first_dt_lot_start,first_dt_lot_last_obs)

### Merging second lot patients to who received ADT in first lot
second_lot <- merge(second_lot, first_adt, by = "subject")

### Now out of this merged cohort choose those who received ARPI within 90 days
second_arpi <- second_lot%>%
  filter(str_detect(treatment_category,"AR Signaling Inhibitor"))%>%
  distinct(subject, .keep_all = T)%>%
  #### Calculating gap
  mutate(gap_second = as.duration(first_dt_lot_start%--%exstdat_int)/ddays(1))%>%
  #### Selecting patients with <90 days gap
  filter(gap_second<=90)

#### Excluding wrongly coded ARPI regimens
second_arpi <- second_arpi%>%
  filter(regimen%in% c("ADT, AR Signaling Inhibitor","ADT, AR Signaling Inhibitor, Chemotherapy","AR Signaling Inhibitor","AR Signaling Inhibitor, NA","AR Signaling Inhibitor, Chemotherapy"))%>%
  rename(second_dt_lot_start = dt_lot_start,
         second_dt_lot_last_obs = dt_lot_last_obs,
         second_regimen = regimen,
         second_treatment_category = treatment_category)

### Renaming columns
second_arpi <- second_arpi%>%
  select(subject,second_regimen,second_treatment_category,treatment_recode)%>%
  rename(treatment_category = second_treatment_category,
         regimen = second_regimen)

########################### Third ARPI

third_lot <- treatment_lot%>%
  filter(lot==3)
### First select who received ADT in first lot
third_lot <- merge(third_lot, first_adt, by = "subject")

### Now out of this merged cohort choose those who received ARPI within 90 days
third_arpi <- third_lot%>%
  filter(str_detect(treatment_category,"AR Signaling Inhibitor"))%>%
  distinct(subject, .keep_all = T)%>%
  mutate(gap_third = as.duration(first_dt_lot_start%--%exstdat_int)/ddays(1))%>%
  filter(gap_third<=90)

### Selecting necessary columns
third_arpi <- third_arpi%>%
  select(subject, regimen, treatment_category, treatment_recode)


### Adding all these to create the cohort of patients who received ARPI
ARPI = dplyr::bind_rows(first_arpi,second_arpi,third_arpi)%>%
  distinct(subject, .keep_all = T)%>%
  mutate(ARPI=1)

```

## Merging date of treatment initiation with the ARPI intensified group
```{r}

### For each patient we selected the earliest date of treatment start (which could be ADT as well) - the reason is even simple ADT can cause PSA to drop

treatment_start_date <- treatment_lot%>%
  group_by(subject)%>%
  slice_min(order_by = dt_lot_start)%>%
  select(subject, dt_lot_start)%>%
  ungroup()%>%
  distinct(subject, .keep_all = T)

ARPI <- left_join(ARPI, treatment_start_date, by = "subject")
```

## Selecting patients with docetaxel
```{r}
### First lot Chemo
first_chemo <- first_lot%>%
  filter(str_detect(treatment_category,"Chemotherapy"))%>%
  distinct(subject, .keep_all = T)


first_chemo <- first_chemo%>%
  select(subject,regimen,treatment_category,treatment_recode)

### Second lot Chemo

second_chemo <- second_lot%>%
  filter(str_detect(treatment_category,"Chemotherapy"))%>%
  distinct(subject, .keep_all = T)%>%
  mutate(gap_second = as.duration(first_dt_lot_start%--%exstdat_int)/ddays(1))%>%
  filter(gap_second<=90)


second_chemo <- second_chemo%>%
  filter(treatment_recode=="Docetaxel")%>%
  rename(second_dt_lot_start = dt_lot_start,
         second_dt_lot_last_obs = dt_lot_last_obs,
         second_regimen = regimen,
         second_treatment_category = treatment_category)


second_chemo <- second_chemo%>%
  select(subject,second_regimen,second_treatment_category,treatment_recode)%>%
  rename(treatment_category = second_treatment_category,
         regimen = second_regimen)

### Third lot Chemo

third_chemo <- third_lot%>%
  filter(str_detect(treatment_category,"Chemotherapy"))%>%
  distinct(subject, .keep_all = T)%>%
  mutate(gap_third = as.duration(first_dt_lot_start%--%exstdat_int)/ddays(1))%>%
  filter(gap_third<=90)

third_chemo <- third_chemo%>%
  select(subject, regimen, treatment_category, treatment_recode)

#### Combining those who received chemo in first, second, or third lot
Chemotherapy <- dplyr::bind_rows(first_chemo,second_chemo,third_chemo)%>%
  distinct(subject, .keep_all = T)%>%
  mutate(Docetaxel=1)
```

## Treatment final coding
```{r}

### Selecting the Subject ID, ARPI, Date of treatment start from ARPI cohort
ARPI <- ARPI%>%
  select(subject, ARPI, dt_lot_start)

### Selecting subject ID from the chemotherapy cohort
Chemotherapy <- Chemotherapy%>%
  select(subject,Docetaxel)

### We will club the ARPI and Docetaxel info with the original cohort
cohort <- left_join(cohort,ARPI, by = "subject")
cohort <- left_join(cohort, Chemotherapy, by = "subject")
cohort$ADT = 1
cohort <- cohort%>%
  replace_na(list(Docetaxel=0, ARPI=0))

###################################################################################
### Treatment recoding to ADT, ADT+Doce, ADT + ARPI, Triplet
cohort$treatment = NA
cohort$treatment[cohort$ARPI==0 & cohort$Docetaxel==0] = "ADT alone"
cohort$treatment[cohort$ARPI==1 & cohort$Docetaxel==0] = "ADT+ARPI"
cohort$treatment[cohort$ARPI==0 & cohort$Docetaxel==1] = "ADT+Docetaxel"
cohort$treatment[cohort$ARPI==1 & cohort$Docetaxel==1] = "Triplet"

#### Selecting patients who received ARPI with ADT or as part of triplet
cohort1 <- cohort%>%
  filter(treatment %in% c("Triplet","ADT+ARPI"))
```

## Merging with baseline data
```{r}
########## Derived Baseline Characteristics
cohort2 <- medidata$subject%>%
  select(subject, is_metastatic_baseline, gleason_bx_mh, gleason_rp_prpros,gleason_bx_prmi,gleason_rp_prmi,gleason_rp_derived,gleason_combined_derived,gleason_factor,origin_metastatic_baseline)

##### Date of consent
consent <- medidata$ic%>%select(subject, cnstdate_int) 

#### adding consent date to the cohort with baseline characteristics
cohort2 <- left_join(cohort2, consent)

####### Merging treatment data (cohort1) with baseline characteristics data
cohort2 <- merge(cohort2, cohort1, by = "subject")

###### Checking for missing data
cohort2%>%
  miss_var_summary()


#### Recoding some of the baseline characteristics data
cohort2 <- cohort2%>%
  mutate(Gleason.total = ifelse(gleason_factor %in% c("0","<=6","7"),"6-7",
                                ifelse(gleason_factor %in% c("8","9","10"),"8-10","Not reported")),
         Liver_mets = ifelse(str_detect(origin_metastatic_baseline,"liver"), "Yes","No"),
         Lung_mets = ifelse(str_detect(origin_metastatic_baseline,"lung"), "Yes","No"),
         Bone_mets = ifelse(str_detect(origin_metastatic_baseline,"thor|spine|Bone|skull|ext|pelvis"),"Yes","No"),
         distal_nodes = ifelse(str_detect(origin_metastatic_baseline,"dn"),"Yes","No"))%>%
  rename(denovomets_yes = is_metastatic_baseline)
```


## Merging Race data
```{r}
Race <- medidata$subject%>%
  select(subject, race_26_1, race_26_2, race_26_3, race_26_4, race_26_5, race_26_6, race_26_7, race_26_other)


Race$Race <- apply(Race[, c('race_26_1', 'race_26_2', 'race_26_3', 'race_26_4', 
                        'race_26_5', 'race_26_6', 'race_26_7', 'race_26_other')], 
                 1, function(x) {
  # Count non-NA values
  non_na_values <- x[!is.na(x)]
  
  # If more than one race is present, assign "Multicategory"
  if (length(non_na_values) > 1) {
    return("Multicategory")
  } else if (length(non_na_values) == 1) {
    return(non_na_values)
  } else {
    return(NA)
  }
})

Race <- Race%>%
  replace_na(list(Race = "Not reported"))

cohort2 <- left_join(cohort2, Race%>%select(subject,Race), by = "subject")
```


## Merging follow-up data
*In 3 cases cause of death was missing. We considered them to die of non-cancer related causes.*
```{r}

### Getting dates from cycles
dates <- medidata$cyc_date


#### Baseline visit date
baseline <- dates%>%
  filter(instance_name=="Baseline 0")%>%
  dplyr::select(subject, visit_date)%>%
  rename(date_baseline = visit_date)

### Selecting the earliest date for each subject
baseline <- baseline%>%
  group_by(subject)%>%
  slice_min(order_by = date_baseline)

baseline <- baseline%>%
  distinct(subject, .keep_all = TRUE)


### Last follow-up date
follow_up <- dates%>%
  group_by(subject)%>%
  slice_max(order_by = visit_date)%>%
  dplyr::select(subject, visit_date)%>%
  rename(date_lfu = visit_date)

follow_up <- follow_up%>%
  distinct(subject, .keep_all = T)

follow_up <- merge(baseline, follow_up, by = "subject")

cohort2 <- merge(cohort2, follow_up, by = "subject")

## Calculating OS from treatment start date
cohort2.1 <- cohort2%>%
  left_join(medidata$pstatus %>% select(subject, deathdate), by = "subject") %>% 
  mutate(OS_date = pmin(date_lfu, deathdate, na.rm = T),
         OS = as.duration(dt_lot_start%--%OS_date)/dmonths(1),
         PCSM = as.factor(ifelse(Dead==0,"Alive",
                       ifelse(Dead==1 & cod=="Prostate Cancer", "PCSM", "OCM"))))

cohort2.1 <- cohort2.1%>%
  replace_na(list(PCSM="OCM")) 
### In 3 cases cause of death was missing. We considered them to die of non-cancer related causes.
###############Univariable Cox regression ##################################################
explan <- c("age","baseline_psa","baseline_psa_cat","alkcat","csliver","cslung","csdn","csskull","csspine","denovomet_disease","treatment","Gleason.total")

cohort2.1%>%
  coxphuni(
    dependent = "Surv(OS,Dead)",
    explanatory = explan
  )%>%
  fit2df()


### Factors associated with OS were age, baseline PSA, baseline ALP, Skull involvement, Gleason score, de novo metastatic presentation, and modest association with lung involvement. 

### Factors associated with PSA nadir were age, baseline PSA, baseline ALP, skull involvement, spine involvement, Gleason score, De novo metastatic disease at presentation. 

### Final factors are ge, baseline PSA, baseline ALP, skull involvement, Gleason score, De novo metastatic disease at presentation
```

## Progression details
```{r}

#### Using newpq version 4
prog0 <- medidata$newdisc_pqv4%>%
  select(subject, instance_name, newpq2_v4, newpq3_v4, newpq4_v4, newpq5_v4, newpq6_v4)%>%
  filter(newpq3_v4=="Yes"|newpq4_v4=="Yes"|newpq5_v4=="Yes"|newpq6_v4=="Yes")%>%
  distinct(subject, .keep_all = T)%>%
  select(subject, instance_name)%>%
  mutate(any_progression = 1)%>%
  mutate(time = gsub("Month", "", instance_name) %>% trimws())%>%
  mutate(time = as.numeric(time))
  
#### Using disc_pq
prog1 <- medidata$disc_pq%>%
  filter(prim_reason=="Disease Progression (Go to C2)")%>%
  select(subject,instance_name,c2_psa1, c2_psa2, c3_psa3,c2_rp1, c2_rp2, c2_rp3,c2_rp4,c2_rp5)%>%
  distinct(subject, .keep_all = T)%>%
  select(subject, instance_name)%>%
  mutate(any_progression = 1)%>%
  mutate(time = gsub("Month", "", instance_name) %>% trimws())%>%
  mutate(time = as.numeric(time))
  
### using new disc_pq version 5
prog2 <- medidata$newdisc_pqv5%>%
  select(subject, instance_name, newpq3_v5, newpq4_v5, newpq5_v5, newpq6_v5)%>%
  mutate(time = gsub("Month", "", instance_name) %>% trimws())%>%
  filter(newpq3_v5=="Yes"|newpq4_v5=="Yes"|newpq6_v5=="Yes"|newpq5_v5=="Yes")%>%
  select(subject, instance_name, time)%>%
  mutate(time = as.numeric(time))%>%
  mutate(any_progression = 1)%>%
  replace_na(list(time=0))


### combining all three and selecting earliest date of progression
prog <- bind_rows(prog0, prog1, prog2)%>%
  group_by(subject)%>%
  slice_min(order_by = time)

#######################################################
dates <- medidata$cyc_date%>%
  select(subject, instance_name, visit_date)

prog <- merge(prog, dates, by = c("subject","instance_name"))%>%
  rename(prog_instance = instance_name,
         prog_time = time, 
         prog_date = visit_date)

cohort2.1 <- left_join(cohort2.1, prog, by = "subject")

cohort2.1 <- cohort2.1%>%
  replace_na(list(any_progression=0))%>%
  mutate(prog_date = pmin(prog_date, OS_date, na.rm = T),
         PFS = as.duration(dt_lot_start%--%prog_date)/dmonths(1),
         cum_prog = ifelse(any_progression==0 & Dead==0,"censored",
                           ifelse(any_progression==1 & PFS<=OS,"progression","competing")),
         progression = ifelse(any_progression==1|Dead==1,1,0))

cohort2.1 <- cohort2.1%>%
  mutate(cum_prog = factor(cum_prog, levels = c("censored","progression","competing")),
         PFS = ifelse(PFS<0,OS,PFS))

######################## Univariable association of progression with baseline factors #########
explan <- c("age","baseline_psa","baseline_psa_cat","alkcat","csliver","cslung","csdn","csskull","csspine","denovomet_disease","treatment","Gleason.total")

cohort2.1%>%
  coxphuni(
    dependent = "Surv(PFS,progression)",
    explanatory = explan
  )%>%
  fit2df() 
### Factors associated with PFS were age, baseline PSA, baseline ALP, Skull involvement, spine involvement, Gleason score, modest association with de novo metastatic presentation. 

### Factors associated with PSA nadir were age, baseline PSA, baseline ALP, skull involvement, spine involvement, Gleason score, De novo metastatic disease at presentation. 

### final factors are ge, baseline PSA, baseline ALP, skull involvement, Gleason score, De novo metastatic disease at presentation, spine involvement

```


```{r}
save(cohort2, file=here("data", "soum_cohort.rdata"))

```

